Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-05-26 tex/manuscript.tex
Lagt inn rettelser

% Running headers of paper:
\markboth%
% First field is the short list of authors
{V. Nygaard, E. A. Rødland, E. Hovig}
% Second field is the short title of the paper
{Exagerated effects after batch correction}

@@

\begin{abstract}
{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous. It also has the option, recommended by the authors \NOTE{Anbefalt av dem alle eller kun av Johnson?}, of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments. Unfortunately, this recommended and frequently used approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner. The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
}
\end{abstract}

@@

The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches. It has thus improved and popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended \CITE{reference}. Based on actual use of ComBat by the authors and others, we suspect thus adjusted data are commonly treated as      ``batch effect free'' in subsequent analysis. And as a consequence, confidence in group effects has been overestimated and false results reported.

@@

The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted "data set 2". Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform. The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch/treatment set up (6-2, 3-4 and 9-6). ComBat was applied followed by a t-test in order to identify differentially expressed genes. First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes (q<0.05).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes (q<0.05). In addition the two sanity checks outline above were performed. In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Figure~\ref{fig:p-johnson} a,b) does not indicate a huge difference. Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis. The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).

>>>>

% Running headers of paper:
\markboth%
% First field is the short list of authors
{V. Nygaard, E. A. Rødland, E. Hovig}
% Second field is the short title of the paper
{Exaggerated effects after batch correction}

@@

\begin{abstract}
{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous. It also has the option of preserving the difference between study groups, estimated from a two-way ANOVA model, to avoid conflating batch effects and group differences during batch adjustments. Unfortunately, this recommended and frequently used approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner. The scientific community seems to be largely unaware of this problem, which most likely has contributed to false discoveries being presented in the published literature.
}
\end{abstract}

@@

The ComBat method described in \citet{Johnson2007}, and included in the sva package (\citealp{Leek2012}), can use either of the two above described approaches to estimate batch differences, but uses an empirical Bayes approach to avoid over-correction for batch effects for small batches. It has thus improved and popularised the two-way ANOVA procedure for retaining group differences when adjusting for batch effects, and the inclusion of group difference as a covariate when removing batch effects has been recommended  in the sva tutorial and user fora. Based on actual use of ComBat by the authors and others, ourselves included, we suspect thus adjusted data are commonly treated as ``batch effect free'' in subsequent analysis. And as a consequence, confidence in group effects has been overestimated and false results reported.

@@

The second example is taken from the supporting information for the original ComBat article (\citealp{Johnson2007}) where it is denoted "data set 2". Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform. The experiment was conducted on three different time points (used as batches) with a total of 30 samples and a fairly balanced batch/treatment set up (6-2, 3-4 and 9-6). ComBat was applied followed by a t-test in order to identify differentially expressed genes. First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes ($q<0.05$).  Then, we analysed their data without batch adjustment in ComBat, but blocking for batch in limma, resulting in 377 probes ($q<0.05$). In addition, the two sanity checks outlined above were performed. In contrast to the above results obtained for(\citealp{Towfic2014}), the P-value distributions for the alternative analyses (Figure~\ref{fig:p-johnson} a,b) does not indicate a huge difference. Nevertheless, we belive that  P-values are deflated for the ComBat adjusted analysis. The R-code for our analysis and a more extensive report can be downloaded from github (\href{https://github.com/vegardny/combat_tests.git}).
